Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-05-05 doc/invt.tex
invt.tex: add the upper bound

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects on the optimal schedule.
%
%This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin (WL) updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
and propose a method of computing
the optimal schedule
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

The matrix $\mathbf w$ is not arbitrary.
%
A necessary condition to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$
is the following.
%
In Eq. \eqref{eq:vt_diffeq_mbin},
if $h_j(t)$ were the same as $p_j$,
the $\dot v_i(t)$ should be a constant independent of $i$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that the elements can be negative.

To simplify the ensuing discussion,
we shall further impose the
detailed balance condition:
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}.
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution $\pmb \rho$
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

@@

Before addressing the optimal schedule
for a multiple-bin scheme,
we shall first compute the equilibrium error
under a constant schedule,
$\alpha(t) = \alpha_0$.
%
After a long period $T'$,
the residual error becomes negligible, and
%
\begin{align}
  \Err
  &=
  \Err_A
  =
  \int_0^{T'}
    \sum_{k = 2}^n
      \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
      e^{ 2 \, \lambda_k \, \alpha_0 \, (t - T') }
    \, dt
  \notag
  \\
  &
  =
  %\stackrel{ T' \to \infty }
  %{ =\joinrel=\joinrel=\joinrel= }
  %
  \sum_{ k = 2 }^n
    \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0
  ,
  \qquad
  ( T' \to \infty )
  .
\notag
%\label{eq:error_eql}
\end{align}
%
In terms of the components, we have
%
\begin{equation}
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0.
  \label{eq:y2_eql}
\end{equation}
%

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved.
%
But the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
coincides with the Lagrangian.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac C T
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$,
%
%We may treat Eq. \eqref{eq:Lagrangian_const}
as a generalization of Eq. \eqref{eq:dudt_const}.
%
By integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $C$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in the above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$
(cf. Sec. \ref{sec:optinitalpha}).
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define the mass function as
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and the normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
Clearly,
both $M(Q)$ and $m(Q)$ decrease monotonically with $Q$.

@@

In each class, the updating magnitudes of two schedules
of lengths $T_1$ and $T_2$
at corresponding times $t_1/T_1 = t_2/T_2$
are related by $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%
Further, by Eqs.
\eqref{eq:error_asym_Lagrangian}
and
\eqref{eq:Lagrangian_const},
we have
%
\begin{equation}
  E_A
  =
  \frac { C^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  ,
\label{eq:error_asym2}
\end{equation}
%
which means the asymptotic error
is inversely proportional to the simulation length
for the same-class schedules,
i.e., we have the scaling relation
$T_1 \, E_{A1} = T_2 \, E_{A2}$.

@@

Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude $\alpha_0$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (\alpha_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial q(T)
  }
  &=
  -\alpha_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial q(T)
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial q(T) = 0$.
}%
By Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we can rewrite the condition as
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ \alpha_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e., the optimal initial updating magnitude
is always half of the equilibration value.
%
This factor $1/2$
happens to coincide with the
reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.

Besides,
for simulations with the same $\alpha_0$,
Eq. \eqref{eq:opt_qT} shows that
the optimal $q(T)$ increases with
the simulation length $T$.
%
If $M(Q)$ decays rapidly,
the distribution $m(Q)$ does not critically
depend on the cutoff $q(T)$,
as long as $q(T)$ is sufficiently large.
%
So the optimal schedules of these simulations
approximately satisfy the scaling relations
described in Sec. \ref{sec:mass_distr}.
%
If, however, $M(Q)$ has a long tail
because of some near-zero $\lambda_k$'s,
increasing the simulation length $T$
generally alters the shape of the optimal schedule.

@@

The Gaussian updating scheme is commonly
used in metadynamics simulations.
%
Below we shall show its stability
in the continuous limit.

@@

We shall first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s,
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
\note{This inequality follows from the non-negativity of
the following quadratic polynomial of $Y$,
$$
\int_0^T
  dt \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
  ,
$$
which implies a non-positive discriminant,
$B^2 - 4 \, A \, C$.
}%
%
Note that the last expression of \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t) \right) = c_k$
for all $k = 2, \dots, n$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
%
\begin{equation}
  \alpha(t) = \frac{              1             }
                   { \lambda_k \, (t + t_{k0} ) },
\label{eq:alpha_invtk}
\end{equation}
with
\begin{equation}
  t_{k0} = \frac{             T            }
                { e^{ \lambda_k q(T) } - 1 }.
\label{eq:tk0}
\end{equation}
%
Such a solution is possible only if
%
\begin{equation}
  \lambda_2 = \dots = \lambda_n.
  \notag
\end{equation}
%
Then, by Eq. \eqref{eq:tk0},
all $t_{k0}$'s also share the same value,
$t_0$,
such that
$u_k = (t + t_0) / (T + t_0)$,
and
$c_k = 1/(T + t_0)$
for $k \ge 2$.
%
The asymptotic error is then
\begin{align}
  E_A
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_singlebin}
\end{align}
%
If we further assume $\lambda_2 = 1$,
%the optimal schedule given by
Eq. \eqref{eq:alpha_invtk}
recovers Eq. \eqref{eq:alpha_invt1}.
%
\note{Derivation.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
    u_k(t) = c_k \left(t + t_{k0} \right).
  \tag{UK}
  \label{neq:ui_solution}
  \end{equation}
  Given that $u_k(T) = 1$ and $u_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{neq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \ln c_k + \ln \left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  Eq. \eqref{eq:alpha_invtk}.
}

Since the first eigenmode corresponds to
a uniform shift of all $v_k$'s,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix
with identical eigenvalues
is a multiple of the identity matrix.
%
Thus,
the single-bin scheme is most efficient
in asymptotic convergence.
%if no eigenvalue $\lambda_k$ is zero.

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
to allow zero eigenvalues.
%
For example,
%if we can assume
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k \le K$,
%
we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
we can change the upper limit of the sum in
Eq. \eqref{eq:error_singlebin} to $K$.
%
%Then,
%Eq. \eqref{eq:error_singlebin} is changed to
%%
%\begin{equation}
%  E_A
%  =
%  \frac {       T     }
%        { (T + t_0)^2 }
%  \sum_{ k = 2 }^K
%    \Gamma_k.
%%\notag
%\label{eq:error_asym_sinc}
%\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time one
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Predicted error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but meet the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

\item
Compute $q(T)$ from the root of Eq. \eqref{eq:opt_qT}.

>>>>

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects on the optimal schedule.
%
%This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

The matrix $\mathbf w$ is not arbitrary.
%
First, to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$,
the $\dot v_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant independent of $i$,
if $h_j(t)$ were the same as $p_j$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that the elements can be negative.

To simplify the ensuing discussion,
we shall limit ourselves to $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution $\pmb \rho$
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

@@

Before addressing the optimal schedule
for a multiple-bin scheme,
we shall first compute the equilibrium error
under a constant schedule,
$\alpha(t) = \alpha_0$.
%
After a long period $T'$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
%\begin{align}
%  \Err
%  &=
%  \Err_A
%  =
%  \int_0^{T'}
%    \sum_{k = 2}^n
%      \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
%      e^{ 2 \, \lambda_k \, \alpha_0 \, (t - T') }
%    \, dt
%  \notag
%  \\
%  &
%  =
%  %\stackrel{ T' \to \infty }
%  %{ =\joinrel=\joinrel=\joinrel= }
%  %
%  \sum_{ k = 2 }^n
%    \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0
%  ,
%  \qquad
%  ( T' \to \infty )
%  .
%\notag
%%\label{eq:error_eql}
%\end{align}
%
%In terms of the components, we have
%
\begin{align}
  \left\langle
    y_k^2
  \right\rangle
  &=
  \int_0^{T'}
    \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
      e^{ 2 \, \lambda_k \, \alpha_0 \, (t - T') }
    \, dt
  \notag
  \\
  &=
  \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0
  ,
  \qquad
  ( T' \to \infty )
  .
\label{eq:y2_eql}
\end{align}
%

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved.
%
But the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
coincides with the Lagrangian.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac C T
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive constant $C$,
%
%We may treat Eq. \eqref{eq:Lagrangian_const}
as a generalization of Eq. \eqref{eq:dudt_const}.
%
By integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $C$ is determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in the above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$
(cf. Sec. \ref{sec:optinitalpha}).
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define the mass function as
\begin{equation}
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
\label{eq:mass_func}
\end{equation}
and the normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%

@@

In each class, the updating magnitudes of two schedules
of lengths $T_1$ and $T_2$
at corresponding times $t_1/T_1 = t_2/T_2$
are related by $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%
Further, by Eqs.
\eqref{eq:error_asym_Lagrangian}
and
\eqref{eq:Lagrangian_const},
we have
%
\begin{equation}
  E_A
  =
  \frac { C^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  ,
\label{eq:error_asym2}
\end{equation}
%
which means the asymptotic error
is inversely proportional to the simulation length
for schedules in the same class,
i.e., we have the scaling relation
$T_1 \, E_{A1} = T_2 \, E_{A2}$.

@@

Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude $\alpha_0$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (\alpha_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial Q_c
  }
  &=
  -\alpha_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial Q_c
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial Q_c = 0$.
}%
By Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we can rewrite the condition as
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ \alpha_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e., the optimal initial updating magnitude
is always half of the equilibration value.
%
This factor $1/2$
happens to coincide with the
reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.

Now consider
simulations with the same $\alpha_0$.
%
Since both $M(Q)$ and $m(Q)$ decrease monotonically with $Q$,
Eq. \eqref{eq:opt_qT} shows that
the optimal $q(T)$ increases with
the simulation length $T$,
more so if $M(Q)$ has a long tail.
%
This means that, if many $\lambda_k$'s are near zero,
both the optimal schedule and
normalized asymptotic error, $E_A \cdot T$,
would be sensitive to the simulation length, $T$.
%
%If $M(Q)$ decays rapidly,
%the distribution $m(Q)$ does not critically
%depend on the cutoff $q(T)$,
%once $q(T)$ is sufficiently large.
%%
%So the optimal schedules of these simulations
%approximately satisfy the scaling relations
%described in Sec. \ref{sec:mass_distr}.
%%
%If, however, $M(Q)$ has a long tail
%because of some near-zero $\lambda_k$'s,
%increasing the simulation length $T$
%generally alters the shape of the optimal schedule.

@@

We define a Gaussian updating scheme
as an updating scheme with a
Gaussian-shaped kernel.
%
This updating scheme is commonly
used in metadynamics.
%
Below we shall show its stability
in the continuous limit.

@@

We shall first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s,
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
\note{This inequality follows from the non-negativity of
the following quadratic polynomial of $Y$,
$$
\int_0^T
  dt \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
  ,
$$
which implies a non-positive discriminant,
$B^2 - 4 \, A \, C$.
}%
%
Note that the last expression of \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t) \right) = c_k$
for all $k = 2, \dots, n$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
%
\begin{equation}
  \alpha(t) = \frac{              1             }
                   { \lambda_k \, (t + t_{k0} ) },
\label{eq:alpha_invtk}
\end{equation}
with
\begin{equation}
  t_{k0} = \frac{             T            }
                { e^{ \lambda_k q(T) } - 1 }.
\label{eq:tk0}
\end{equation}
%
Such a solution is possible only if
%
\begin{equation}
  \lambda_2 = \dots = \lambda_n.
  \notag
\end{equation}
%
Then, by Eq. \eqref{eq:tk0},
all $t_{k0}$'s also share the same value,
$t_0$,
such that
$u_k = (t + t_0) / (T + t_0)$,
and
$c_k = 1/(T + t_0)$
for $k \ge 2$.
%
The asymptotic error is then
\begin{align}
  E_A
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_asym_singlebin}
\end{align}
%
If we further assume $\lambda_2 = 1$,
%the optimal schedule given by
Eq. \eqref{eq:alpha_invtk}
recovers Eq. \eqref{eq:alpha_invt1}.
%
\note{Derivation.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
    u_k(t) = c_k \left(t + t_{k0} \right).
  \tag{UK}
  \label{neq:ui_solution}
  \end{equation}
  Given that $u_k(T) = 1$ and $u_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{neq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \ln c_k + \ln \left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  Eq. \eqref{eq:alpha_invtk}.
}

Since the first eigenmode corresponds to
a uniform shift of all $v_k$'s,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix
with identical eigenvalues
is a multiple of the identity matrix.
%
Thus,
the corresponding updating scheme,
the single-bin scheme, is most efficient
in asymptotic convergence.
%if no eigenvalue $\lambda_k$ is zero.
%
In Appendix \ref{sec:upperbound},
we shall show that the opposite limit of
a set of well-separated eigenvalues
will provide an upper bound for
the asymptotic error, Eq. \eqref{eq:error_asym_ubound}.

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
to allow zero eigenvalues.
%
For example,
%if we can assume
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k \le K$,
%
we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
we can change the upper limit of the sum in
Eq. \eqref{eq:error_asym_singlebin} to $K$:
%
%Then,
%Eq. \eqref{eq:error_singlebin} is changed to
%
\begin{equation}
  E_A
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
%\notag
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time one
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.
%
The definition of $K$ will be modified.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Predicted error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $k \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the relatively flat profile,
    but meet the initial error curve
    at a lower level (not shown),
    where $\lambda_k \approx 1/q(T)$
    (cf. Appendix \ref{sec:upperbound}).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

\item
Compute the optimal $q(T)$ by solving Eq. \eqref{eq:opt_qT}.
